Appendix 3: Estimating species interactions

### 2015-10-31 Harris_2015/manuscript-materials/draft.tex
closer

\documentclass[12pt,]{article}
\usepackage[compact]{titlesec}
\usepackage{fancyhdr}
\usepackage{hanging}
\pagestyle{fancy}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\usepackage{ifthen}
\geometry{verbose,letterpaper,tmargin=2.35cm, bmargin=2.35cm,lmargin=2.44cm,rmargin=2.44cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\setlength{\parindent}{1cm}

To the extent that nontrophic species interactions (such as competition)
affect community assembly, ecologists might expect to find signatures of
these interactions in species composition data (MacArthur 1958, Diamond
1975). Despite decades of work and several major controversies, however
(Lewin 1983, Strong et al. 1984, Connor et al. 2013), existing methods
for detecting competition's effects on community structure are
unreliable (Gotelli and Ulrich 2009). In particular, species' effects on
one another can become lost in the complex web of indirect effects. For
example, the competitive interaction between the two shrub species in
Figure 1A can become obscured by their shared tendency to occur in
unshaded areas (Figure 1B). While ecologists have long known that
indirect effects can overwhelm direct ones at the landscape level
(Dodson 1970, Levine 1976), the vast majority of our methods for drawing
inferences from observational data do not control for these effects
(e.g. Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech
2013, Pollock et al. 2014). To the extent that indirect interactions
like those in Figure 1 are generally important (Dodson 1970), existing
methods will thus not provide much evidence regarding species
interactions.

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most methods assume), it nevertheless does
leave a signal in the data (Figure 1C). For example, after partitioning
the data set into shaded sites and unshaded sites, there will be
co-occurrence deficits in each subset that might not be apparent at the
whole-landscape level. More generally, controlling for other species in
the network will often be important for obtaining reliable estimates of
direct (conditional, or all-else-equal) effects. This kind of precision
is difficult to obtain from null models, which only simulate the most
extreme possible scenario: that \emph{all} direct and indirect
interactions are exactly zero. Nevertheless, null models have dominated
this field for more than three decades (Strong et al. 1984, Gotelli and
Ulrich 2009).

Following Azaele et al. (2010), this paper shows that Markov networks
(undirected graphical models also known as Markov random fields; Murphy
2012) can provide a framework for understanding the landscape-level
consequences of pairwise species interactions, and for detecting them
from observed presence-absence matrices. Markov networks have been used
in many scientific fields in similar contexts for decades, from physics
(where nearby particles interact magnetically; Cipra 1987) to spatial
statistics (where adjacent grid cells have correlated values; Harris
1974, Gelfand et al. 2005). While community ecologists explored some
related approaches in the 1980's (Whittam and Siegel-Causey 1981), they
used severe approximations that led to unintelligible results (e.g.
``probabilities'' greater than one; Gilpin and Diamond 1982).

Below, I introduce Markov networks and show how they can be used to
simulate landscape-level data from an interaction matrix or to make
exact predictions about the direct and indirect consequences of possible
interaction matrices. Then, using simulated data sets where the ``true''
interactions are known, I compare this approach with several existing
methods. Finally, I discuss opportunities for extending the approach
presented here to other problems in community ecology, e.g.~quantifying
the overall effect of species interactions on occurrence rates
(Roughgarden 1983) and disentangling the effects of biotic versus
abiotic interactions on species composition (Pollock et al. 2014).

@@

\raggedright
\setlength{\parindent}{1cm}

\noindent where the second sum is over all \(\frac{1}{2}n(n-1)\) pairs
of \(n\) species. In this model, \(\alpha_{i}\) is an intercept term
determining the amount that the presence of species \(i\) contributes to
the log-probability of \(\vec{y}\); it directly controls the prevalence
of species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
co-occurrence of species \(i\) and species \(j\) contributes to the
log-probability; it controls the conditional relationship between two
species, i.e.~the probability that they will be found together, after
controlling for the other species in the network (Figure 2A, Figure 2B).
For example, \(\beta_{ij}\) might have a value of \(+2\) for two
mutualists, indicating that the odds of observing one species are
\(e^2\) times higher in sites where its partner is present than in
comparable sites where its partner is absent.

Because the relative probability of a presence-absence vector increases
when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends---all else
equal---to produce assemblages that have many pairs of
positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect). Of
course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then species' overall co-occurrence patterns can be more
complicated, and may require summing over the different possible
assemblages (Figure 2B) to determine the overall outcomes.

@@

\noindent \textbf{Evaluating model performance.} For the simulated
landscapes based on Figure 1, I assessed whether each method's test
statistic indicated a positive or negative relationship between the two
shrubs (Appendix 1). For the null model (Pairs), I calculated
statistical significance using its \(Z\)-score. For the Markov network,
I used the Hessian matrix to generate approximate confidence intervals
and noted whether these intervals included zero.

I then evaluated the relationship between each method's estimates and
the ``true'' interaction strengths among all of the species pairs from
the larger simulated landscapes. This determined which of the methods
provide a consistent way to know how strong species interactions
are---regardless of which species were present in a particular data set
or how many observations were taken. To ensure that the different test
statistics (e.g.~correlations versus \(Z\) scores) were on a common
scale, I rescaled them using linear regression through the origin. After
rescaling each method's estimates, I calculated squared errors between
the scaled interaction estimates and ``true'' interaction values across
the simulated data sets. These squared errors determined the proportion
of variance explained for different combinations of model type and
landscape size (compared with a null model that assumed all interaction
strengths to be zero).

@@

\noindent
\textbf{Twenty species.} Despite some variability across contexts
(Figure 3), the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) outperformed those that did not. Of these,
the Markov network consistently performed the best, especially on the
smaller landscapes (which are also the most representative of typical
analyses in this field; Gotelli and Ulrich 2009). As shown in Appendix
{[}{[}X{]}{]}, the standard errors associated with the estimates in
Figure 3 are all small (less than 0.01), so differences across methods
should not be attributed to sampling error.

Of the methods that did not control for indirect effects, Figure 3 shows
that simple correlation coefficients provided a more reliable indicator
of species' true interaction strengths than either the joint species
distribution model (BayesComm) or the null model (Pairs). The estimates
from these three approaches were tightly correlated (Figure 3B),
suggesting that the effect sizes from the null model (and, to a lesser
extent, the correlation matrices from joint species distribution models)
only contain noisy versions of the same information that could be
obtained more easily and interpretably by calculating correlation
coefficients between species' presence-absence vectors.

Whereas Figure 3 summarizes the accuracy of each model's point
estimates, Figure 4 compares the inferential value of the Markov network
versus the null model. When the ``true'' \(\beta\) coefficient is near
zero, the Markov network's 95\% confidence intervals usually include
zero (but not quite 95\% of the time), and it is rare for the model to
confidently attribute the wrong sign to a coefficient (Figure 4A). The
null model, by contrast, typically rejects the null hypothesis at the
\(p=0.05\) level regardless of the underlying parameters, and the sign
of its \(Z\)-scores is not a reliable indicator of the interaction's
actual sign.

@@

The results presented above show that Markov networks can reliably
recover species' pairwise interactions from observational data, even for
cases where environmental heterogeneity and indirect interactions cause
ecologists' typical null modeling approaches to reliably fail. For cases
where fitting a Markov network is computationally infeasible, these
results also indicate that partial covariances and generalized linear
models can both provide useful approximations. The partial correlations'
success on simulated data may not carry over to real data sets, however;
Loh and Wainwright (2013) show that the linear approximations can be
less reliable in cases where the true interaction matrix contains more
structure (e.g.~guilds or trophic levels). Similarly, the approximation
involved in using separate generalized linear models for each species
can occasionally lead to catastrophic overfitting with small-to-moderate
sample sizes (Lee and Hastie 2012). For these reasons, it will usually
be best to fit a Markov network rather than one of the alternative
methods when one's computational resources allow it.

It's important to note that none of these methods can identify the exact
nature of the pairwise interactions (e.g.~which species in a
positively-associated pair is facilitating the other; Schmidt and Murphy
2012), particularly when real pairs of species can reciprocally
influence one another in multiple ways simultaneously (Bruno et al.
2003); with compositional data, there is only enough information to
provide a single number describing each species pair. To estimate
asymmetric interactions, such as commensalism or predation, ecologists
would need other kinds of data, as from time series, behavioral
observations, manipulative experiments, or natural history. These other
sources of information could also be used to augment the likelihood
function with an informative prior distribution, which could lead to
better results on some real data sets than was shown in Figure 3A.

Despite their limitations, Markov networks have enormous potential to
improve ecological understanding. In particular, they are less
vulnerable than some of the most commonly-used methods to mistakenly
identifying positive species interactions between competing species, and
can make precise statements about the conditions where indirect
interactions will overwhelm direct ones. They also provide a simple
answer to the question of how competition should affect a species'
overall prevalence, which was a major flashpoint for the null model
debates in the 1980's (Roughgarden 1983, Strong et al. 1984). Equation 1
can be used to calculate the expected prevalence of a species in the
absence of biotic influences (\(\frac{e^\alpha}{1 + e^{\alpha}}\); Lee
and Hastie 2012). Competition's effect on prevalence in a Markov network
can then be calculated by subtracting this value from the observed
prevalence (cf Figure 2D). This kind of insight would have been
difficult to obtain without a generative model that makes predictions
about the consequences of species interactions; null models (which
presume \emph{a priori} that interactions do not exist) have no way to
make such predictions.

Markov networks---particularly the Ising model for binary
networks---have been studied for nearly a century (Cipra 1987), and the
models' properties, capabilities, and limits are well-understood. Using
the same framework for species interactions would thus allow ecologists
to tap into an enormous set of existing discoveries and techniques for
dealing with indirect effects, stability, and alternative stable states.
Numerous other extensions are possible: for example, the states of the
interaction network can be modeled as a function of the local abiotic
environment (Lee and Hastie 2012), which would provide a rigorous and
straightforward approach to the difficult and important task of
incorporating whole networks of biotic interactions into species
distribution models (Pollock et al. 2014), leading to a better
understanding of the interplay between biotic and abiotic effects on
community structure. Alternatively, models could allow one species to
affect the interactions between two other species (Tjelmeland and Besag
1998), tipping the balance between facilitation and exploitation (Bruno
et al. 2003).

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species
interactions. Null and neutral models can be useful for clarifying our
thinking about the numerical consequences of species' richness and
abundance patterns (Harris et al. 2011, Xiao et al. 2015), but
deviations from a particular null model must be interpreted with care
(Roughgarden 1983). Even in small networks with three species, it may
simply not be possible to implicate individual species pairs or specific
ecological processes like competition by rejecting a general-purpose
null (Gotelli and Ulrich 2009), especially when the test statistic is
effectively just a correlation coefficient {[}{[}Figure 3B{]}{]}. When a
non-null backdrop is not controlled for, Type I error rates can
skyrocket, the apparent sign of the interaction can change, and null
models can lead to reliably incorrect inferences (Figure 1, Figure 4).

@@

\noindent \textbf{Figure 2: A.} A small Markov network with two species,
defined by its \(\alpha\) and \(\beta\) values. The abiotic environment
favors the occurrence of each species (\(\alpha >0\)), particularly
species 2 (\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient
linking these two species implies that they co-occur less than expected
under independence. \textbf{B.} These coefficients determine the
probabilities of all four possible presence-absence combinations for
Species 1 and Species 2. \(\alpha_1\) is added to the exponent whenever
Species 1 is present (\(y_1 = 1\)), but not when it is absent
(\(y_1 = 0\)). Similarly, the exponent includes \(\alpha_2\) only when
species \(2\) is present (\(y_2 = 1\)), and \(\beta\) only when both are
present (\(y_1y_2 = 1\)). The normalizing constant \(Z\), ensures that
the four probabilities sum to 1. In this case, \(Z\) is about 18.5.
\textbf{C.} The expected frequencies of all possible co-occurrence
patterns between the two species of interest, as calculated in the
previous panel. \textbf{D.} If, as assumed in null model simulations,
\(\beta_{12}\) equaled zero (e.g.~if the species no longer competed for
the same resources), then the reduction in competition would allow each
species to increase its overall occurrence rate.

\noindent \textbf{Figure 3: A.} Proportion of variance in interaction
coefficients explained by each method versus number of sampled
locations. \textbf{B.} The \(Z\)-scores produced by the null model
(``Pairs'') for each pair of species can be predicted using the
correlation between the presence-absence vectors of those same species
and from the number of sites on the landscape.

>>>>

\documentclass[12pt,]{article}
\usepackage[compact]{titlesec}
\usepackage{fancyhdr}
\usepackage{hanging}
\pagestyle{fancy}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\usepackage{ifthen}
\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\setlength{\parindent}{2em}

To the extent that nontrophic species interactions (such as competition)
affect community assembly, ecologists might expect to find signatures of
these interactions in species composition data (MacArthur 1958, Diamond
1975). Despite decades of work and several major controversies, however
(Lewin 1983, Strong et al. 1984, Connor et al. 2013), existing methods
for detecting competition's effects on community structure are
unreliable (Gotelli and Ulrich 2009). In particular, species' effects on
one another can become lost in a web of indirect effects. For example,
the competitive interaction between the two shrub species in Figure 1A
is obscured by their shared tendency to occur in unshaded areas (Figure
1B). While ecologists have long known that indirect effects can
overwhelm direct ones at the landscape level (Dodson 1970, Levine 1976),
the vast majority of our methods for drawing inferences from
observational data do not control for these effects (e.g. Diamond 1975,
Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013, Pollock et al.
2014). To the extent that indirect interactions like those in Figure 1
are generally important (Dodson 1970), existing methods will not provide
much evidence regarding species interactions.

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most methods assume), it nevertheless does
leave a signal in the data (Figure 1C). For example, after partitioning
the data set into shaded and unshaded sites, there will be co-occurrence
deficits in each subset that wouldn't otherwise be apparent. More
generally, controlling for other species in the network will often be
important for obtaining reliable estimates of direct (conditional, or
all-else-equal) effects. This kind of precision is difficult to obtain
from null models, which only simulate the most extreme possible
scenario: that \emph{all} direct and indirect interactions are exactly
zero. Nevertheless, null models have dominated this field for more than
three decades (Strong et al. 1984, Gotelli and Ulrich 2009).

Following Azaele et al. (2010), this paper shows that Markov networks
(undirected graphical models also known as Markov random fields; Murphy
2012) can provide a framework for understanding the landscape-level
consequences of pairwise species interactions, and for estimating them
from observed presence-absence matrices. Markov networks have been used
in many scientific fields in similar contexts for decades, from physics
(where nearby particles interact magnetically; Cipra 1987) to spatial
statistics (where adjacent grid cells have correlated values; Harris
1974, Gelfand et al. 2005). While community ecologists explored some
related approaches in the 1980's (Whittam and Siegel-Causey 1981), they
used severe approximations that led to unintelligible results (e.g.
``probabilities'' greater than one; Gilpin and Diamond 1982).

Below, I demonstrate Markov networks' ability to produce exact
predictions about the direct and indirect consequences of an interaction
matrix, and also to make inferences about the species interactions that
contributed to an observed set of co-occurrences. Using simulated data
sets where the ``true'' interactions are known, I compare this approach
with several existing methods. Finally, I discuss opportunities for
extending the approach presented here to other problems in community
ecology, e.g.~quantifying the overall effect of species interactions on
occurrence rates (Roughgarden 1983) and disentangling the effects of
biotic versus abiotic interactions on species composition (Pollock et
al. 2014).

@@

\raggedright
\setlength{\parindent}{2em}

\noindent where the second sum is over all \(\frac{1}{2}n(n-1)\) pairs
of \(n\) species. In this model, \(\alpha_{i}\) is an intercept term
determining the amount that the presence of species \(i\) contributes to
the log-probability of \(\vec{y}\); it directly controls the prevalence
of species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
co-occurrence of species \(i\) and species \(j\) contributes to the
log-probability; it controls the conditional relationship between two
species, i.e.~the probability that they will be found together, after
controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if \(\beta_{ij} = -2\), then each species' odds of
occurrence are \(exp{2}\) times lower when the other one is present (as
compared with otherwise comparable sites). Because the relative
probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends---all else
equal---to produce assemblages that have many pairs of
positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect). Of
course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then species' overall co-occurrence patterns can be more
complicated, and may require summing over the different possible
assemblages (Figure 2B) to determine the overall outcomes.

@@

\noindent \textbf{Evaluating model performance.} For the simulated
landscapes based on Figure 1, I assessed whether each method's test
statistic indicated a positive or negative relationship between the two
shrubs (Appendix 1). For the null model (Pairs), I calculated
statistical significance using its \(Z\)-score. For the Markov network,
I used the Hessian matrix to generate approximate confidence intervals.

I then evaluated the relationship between each method's estimates and
the ``true'' interaction strengths among all of the species pairs from
the larger simulated landscapes. This determined which of the methods
provide a consistent way to know how strong species interactions
are---regardless of which species were present in a particular data set
or how many observations were taken. To ensure that the different test
statistics (e.g.~correlations versus \(Z\) scores) were on a common
scale, I rescaled them using linear regression through the origin. I
then calculated the proportion of variance explained for different
combinations of model type and landscape size (compared with a null
model that assumed all interaction strengths to be zero).

@@

\noindent
\textbf{Twenty species.} Despite some variability across contexts
(Figure 3), the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) outperformed those that did not. Of these,
the Markov network consistently performed the best, especially on the
smaller landscapes (which are also the most representative of typical
analyses in this field; Gotelli and Ulrich 2009). As shown in Appendix
{[}{[}X{]}{]}, the standard errors associated with the estimates in
Figure 3 are small (less than 0.01), so these differences should not be
attributed to sampling error.

Of the methods that did not control for indirect effects, Figure 3 shows
that simple correlation coefficients provided a more reliable indicator
of species' true interaction strengths than either the joint species
distribution model (BayesComm) or the null model (Pairs). The estimates
from these three approaches were tightly correlated (Figure
{[}{[}3B{]}{]}), suggesting that the null model only contains a noisy
version of the same information that could be obtained more easily and
interpretably with simple correlation coefficients.

Finally, we can evaluate the models' statistical inferences. Figure 4
shows, across a range of true interaction strengths, the probability
that the null model or the Markov network will predict the wrong sign of
the interaction with 95\% confidence. The null model makes such errors
nearly 8 times as often as the Markov network, even though it only
reject the null hypothesis twice as often overall {[}{[}Appendix{]}{]}.
The Markov network's errors were also more concentrated around 0, so it
never misclassified strong interactions like the null model did.

@@

The results presented above show that Markov networks can reliably
recover species' pairwise interactions from observational data, even for
cases where environmental heterogeneity and indirect interactions cause
ecologists' typical null modeling approaches to reliably fail. Partial
covariances and generalized linear models can both provide
computationally efficient approximations, but with somewhat lower
accuracy. The difference in accuracy may be larger for real data sets
than for the simulated landscapes in Figure 3, however; linear
approximations to the Markov network make larger errors when the
interaction matrix is structured (e.g.~due to guilds or trophic levels;
Loh and Wainwright 2013). Similarly, the separate generalized linear
models for each species can severely overfit in some cases (Lee and
Hastie 2012). The full Markov network should thus be preferred to the
approximations when it is computationally feasible.

Compositional data only has enough degrees of freedom to estimate one
interaction per species pair (Schmidt and Murphy 2012), so none of these
methods can identify the exact nature of the pairwise interactions
{[}e.g.~which species in a positively-associated pair is facilitating
the other{]}. To estimate asymmetric interactions, such as commensalism
or predation, ecologists could use time series, behavioral observations,
manipulative experiments, or natural history. These other sources of
information could also be used to augment the likelihood function with
an informative prior distribution, reducing ecologists' eror and
uncertainty.

Despite their limitations, Markov networks have enormous potential to
improve ecological understanding. In particular, they are less
vulnerable than some of the most commonly-used methods to mistakenly
identifying positive species interactions between competing species, and
can make precise statements about the conditions where indirect
interactions will overwhelm direct ones. They also provide a simple
answer to the question of how competition should affect a species'
overall prevalence, which was a major flashpoint for the null model
debates in the 1980's (Roughgarden 1983, Strong et al. 1984). Equation 1
can be used to calculate the expected prevalence of a species in the
absence of biotic influences (\(\frac{e^\alpha}{1 + e^{\alpha}}\); Lee
and Hastie 2012). Competition's effect on prevalence in a Markov network
can then be calculated by subtracting this value from the observed
prevalence (cf Figure 2D). This kind of insight would have been
difficult to obtain without a generative model that predicts the
consequences of species interactions (as opposed to null simulations
that omit interactions entirely).

Markov networks---particularly the Ising model for binary
networks---have been studied for nearly a century (Cipra 1987), and the
models are well-understood. Using the same framework for species
interactions would thus allow ecologists to tap into a vast set of
existing discoveries and techniques for dealing with indirect effects,
stability, and alternative stable states. Numerous other extensions are
possible: for example, the states of the interaction network can be
modeled as a function of the local abiotic environment (Lee and Hastie
2012), which would help incorporate networks of biotic interactions into
species distribution models (Pollock et al. 2014) and lead to a better
understanding of the interplay between biotic and abiotic effects on
community structure. Alternatively, models could allow one species to
alter the relationship between two other species (Tjelmeland and Besag
1998, cf Bruno et al. 2003).

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species
interactions. Null and neutral models can be useful for clarifying our
thinking (Harris et al. 2011, Xiao et al. 2015), but deviations from a
particular null model must be interpreted with care (Roughgarden 1983).
Even in small networks with three species, it may simply not be possible
to implicate specific ecological processes like competition by rejecting
a general-purpose null (Gotelli and Ulrich 2009), especially when the
test statistic is effectively just a correlation coefficient
{[}{[}Figure 3B{]}{]}. When a non-null backdrop is not controlled for,
Type I error rates can skyrocket, the apparent sign of the interaction
can change, and null models can lead to reliably incorrect inferences
(Figure 1, Figure 4).

@@

\noindent \textbf{Figure 2: A.} A small Markov network, defined by its
\(\alpha\) and \(\beta\) values. The abiotic environment favors the
occurrence of each species (\(\alpha >0\)), particularly species 2
(\(\alpha_2 > \alpha_1\)). The negative \(\beta_12\) coefficient is
consistent with competition between the two species. \textbf{B.} The
coefficients determine the probabilities of all four possible
presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
and includes \(\beta_12\) only when both are present (\(y_1y_2 = 1\)).
The normalizing constant \(Z\), ensures that the four probabilities sum
to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
frequencies of all possible co-occurrence patterns between the two
species of interest, as calculated in the previous panel. \textbf{D.} If
\(\beta_{12}\) equaled zero (e.g.~if the species no longer competed for
the same resources), then the each species' overall occurrence rates
would increase.

\noindent \textbf{Figure 3:} Proportion of variance in interaction
coefficients explained by each method versus number of sampled
locations.
